\documentclass{article}
\input{../../../../lib.sty}

\title{\texttt{MakeNoise<MapDomain<AtomDomain<TK>, AtomDomain<TV>>, L0PInfDistance<P, AbsoluteDistance<QI>>, MO> for IntExpFamily<P>}}
\author{Michael Shoemate}
\date{}

\begin{document}

\maketitle

\contrib
Proves soundness of the implementation of \rustdoc{measurements/noise_threshold/trait}{MakeNoiseThreshold} over hashmaps
for \texttt{IntExpFamily} in \asOfCommit{mod.rs}{f5bb719}.

This mechanism samples from the \texttt{IntExpFamily} distribution, 
where the tails are clipped to the smallest and largest representable integers in the native integer type.
This is done by first sampling from the \texttt{ZExpFamily}, the equivalent distribution supported on all integers,
and then clipping the tails by clamping.
The clamping is done by saturating the cast of the sampled value to the native integer type.

\section{Hoare Triple}
\subsection*{Precondition}
\subsubsection*{Compiler-Verified}

\begin{itemize}
    \item Generic \texttt{T} implements trait \rustdoc{traits/trait}{Integer} and \rustdoc{traits/trait}{SaturatingCast}\texttt{<IBig>}
        The saturating cast is for infallible postprocessing of big ints back to type \texttt{T}.
    \item Const-generic \texttt{P} is of type \texttt{usize}
    \item Generic \texttt{QI} implements trait \rustdoc{traits/trait}{Integer}
    \item Generic \texttt{MO} implements trait \rustdoc{core/trait}{Measure}
    \item Type \texttt{IBig} implements trait \texttt{From<T>}.
        This infallible exact cast is for converting integers to big ints in the preprocessing transformation.
    \item Type \texttt{RBig} implements trait \texttt{TryFrom<QI>}.
        This is for fallible casting from input sensitivity of type \texttt{QI} to a rational in the privacy map.
    \item Type \texttt{ZExpFamily<P>} implements trait \rustdoc{measurements/trait}{NoisePrivacyMap}\texttt{<LpDistance<P, RBig>, MO>}.
        This bound requires that it must be possible to construct a privacy map for this combination of noise distribution, distance type and privacy measure.
\end{itemize}

\subsubsection*{User-Verified}
None

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/MakeNoiseThreshold_MapDomain_for_IntExpFamily.py}

\subsection*{Postcondition}
\begin{theorem}
    \validMeasurement{(\texttt{self, input\_space, threshold, MO, TK, TV, P, QI})}{\texttt{make\_noise\_threshold}}
\end{theorem}

\begin{proof}
    Line \ref{line:dist} constructs a new random variable following a distribution equivalent to \texttt{IntExpFamily}, 
    but without clipped tails.

    Neither constructor \rustdoc{measurements/noise_threshold/nature/integer/fn}{make\_int\_to\_bigint\_threshold} nor
    \rustdoc{measurements/noise_threshold/trait}{MakeNoiseThreshold}\texttt{.make\_noise\_threshold} have manual preconditions, 
    and the postconditions guarantee a valid transformation and valid measurement, respectively.
    \rustdoc{measurements/noise_threshold/nature/integer/fn}{then\_saturating\_cast\_hashmap} also does not have preconditions,
    and its postcondition guarantees that it returns a valid postprocessor.

    The chain of a valid transformation, valid measurement and valid postprocessor is a valid measurement.
\end{proof}

\end{document}
