\documentclass{article}
\input{../../../../lib.sty}



\title{\texttt{fn make\_scalar\_integer\_laplace}}
\author{Michael Shoemate}
\begin{document}

\maketitle

\contrib

Proves soundness of \rustdoc{measurements/fn}{make\_scalar\_integer\_laplace} in \asOfCommit{mod.rs}{f5bb719}.
The function on the resulting measurement takes in a data set \texttt{x} (a single integer), 
and returns a sample from the Discrete Laplace Distribution centered at \texttt{x}, with a fixed noise scale.

\subsection*{PR History}
\begin{itemize}
    \item \vettingPR{490}
\end{itemize}

\section{Hoare Triple}

\subsection*{Preconditions}
\begin{itemize}
    \item Variable \texttt{input\_domain}, of type \texttt{AtomDomain<T>}
    \item Variable \texttt{input\_metric}, of type \texttt{AbsoluteDistance<T>}
    \item Variable \texttt{scale}, of type \texttt{QO}
    \item Type \texttt{T} must have trait \rustdoc{traits/trait}{Integer} and support saturating cast from \texttt{IBig} (for postprocessing a noisy big integer back to \texttt{T})
    \item Type \texttt{QO} must have trait \rustdoc{traits/trait}{Float} and support casting with controlled rounding from \texttt{T} (for converting $d_{in}$ to type \texttt{QO})
    \item Type \texttt{IBig} must be constructable from \texttt{T} (to convert the data into a big integer)
    \item Type \texttt{RBig} must be fallibly constructable from \texttt{QO} (to convert scale into a rational)
\end{itemize}

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/make_scalar_integer_laplace.py}

\subsection*{Postcondition}

\validMeasurement{\texttt{(input\_domain, input\_metric, scale, T, QO)}}{\\\texttt{make\_scalar\_integer\_laplace}}

\section{Proof}

\begin{proof} 
\textbf{(Privacy guarantee.)} 
    
\begin{tcolorbox}
    The proof assumes the following lemma.
    \begin{lemma}
        \texttt{sample\_integer\_laplace} and \texttt{laplace\_map} each satisfy their postcondition.
    \end{lemma}
\end{tcolorbox}

\texttt{sample\_integer\_laplace} can only fail due to lack of system entropy. 
This is usually related to the computer's physical environment and not the dataset. 
The rest of this proof is conditioned on the assumption that this function does not raise an exception. 

Let $x$ and $x'$ be datasets that are \texttt{d\_in}-close with respect to \texttt{input\_metric}.
Here, the metric is \texttt{AbsoluteDistance<T>}.

By the postcondition of \texttt{sample\_integer\_laplace},
the output of the function follows the Discrete Laplace Distribution with scale \texttt{scale}.

\begin{align*}
    & \max_{x \sim x'} D_{\infty}(M(x), M(x'))  \\
    =& \max_{x \sim x'} \max_{z \in supp(M(\cdot))} \ln\left(\frac{\Pr\left[M(x) = z\right]}{\Pr\left[M(x') = z \right]}\right)
        &&\text{substitute } D_{\infty}\\
    =& \max_{x \sim x'} \max_{z \in \mathbb{Z}} \ln\left(\frac{\Pr\left[\mathrm{DLap}(x, b) = z \right]}{\Pr\left[\mathrm{DLap}(x', b) = z\right]}\right)
        &&\text{where } b \text{ is the noise scale} \\
    =& \max_{x \sim x'} \max_{z \in \mathbb{Z}} \ln\left(\frac{
    \frac{\exp^{1/b} - 1}{\exp^{1/b} + 1} \exp \left( -\frac{|x - z|}{b} \right)
    }{
        \frac{\exp^{1/b} - 1}{\exp^{1/b} + 1} \exp \left( -\frac{|x' - z|}{b} \right)
    }\right) 
        &&\text{use pdf of Discrete Laplace} \\
    =& \max_{x \sim x'} \max_{z \in \mathbb{Z}} \ln\left(\frac{
    \exp \left( -\frac{|x - z|}{b} \right)
    }{
        \exp \left( -\frac{|x' - z|}{b} \right)
    }\right) \\
    =& \max_{x \sim x'} \max_{z \in \mathbb{Z}}\frac{|x' - z| - |x - z|}{b}&& \text{exp and ln cancel} \\
    \leq& \frac{\max_{x \sim x'} |x - x'|}{b} &&\text{by reverse triangle inequality} \\
    =& \frac{d_{in}}{b}  &&\text{by definition of absolute distance}
\end{align*}

This bound satisfies the postcondition of \texttt{laplace\_map}.
The saturating conversion to \texttt{T} is a post-processing step.

Therefore it has been shown that for every pair of elements $x, x' \in \texttt{input\_domain}$ and every $d_{Abs}(x, x') \le \din$ with $\din \ge 0$, 
if $x, x'$ are $\din$-close then $\function(x),\function(x')$ are $\texttt{privacy\_map}(\din)$-close under $\texttt{output\_measure}$ (the Max-Divergence).
\end{proof}

\end{document}
