\documentclass{article}
\input{../../../../lib.sty}

\title{\texttt{MakeNoiseThreshold<DI, MI, MO> for DiscreteLaplace}}
\author{Michael Shoemate}
\date{}

\begin{document}

\maketitle

\contrib
Proves soundness of the implementation of \rustdoc{measurements/noise_threshold/trait}{MakeNoiseThreshold} 
for \texttt{DiscreteLaplace} in \asOfCommit{mod.rs}{f5bb719}.

This is an intermediary compile-time layer whose purpose is 
to dispatch to either the integer or floating-point variations of the mechanism,
depending on the type of data in the input domain.

It does this through the use of the \rustdoc{measurements/noise/nature/trait}{Nature} trait,
which has concrete implementations for each possible input type.
This layer makes interior layers simpler to work with, and does not have privacy implications.
It also makes \texttt{make\_laplace\_threshold} easier to call,
by simplifying the type signature.

\section{Hoare Triple}
\subsection*{Precondition}
\subsubsection*{Compiler-Verified}
\texttt{MakeNoiseThreshold} is parameterized as follows:
\begin{itemize}
    \item \texttt{DI} implements trait \rustdoc{core/trait}{Domain}
    \item \texttt{MI} implements trait \rustdoc{core/trait}{Metric}
    \item \texttt{MO} implements trait \rustdoc{core/trait}{Measure}
\end{itemize}

The following trait bounds are also required:
\begin{itemize}
    \item \texttt{(DI, MI)} implements trait \texttt{MetricSpace}
    \item \texttt{DI\_Atom} implements trait \rustdoc{measurements/noise/nature/trait}{Nature}.
        This trait encodes the relationship between the atomic data type 
        and the type of the noise distribution that is compatible with it: \texttt{DI\_Atom\_RV2}.
        In Rust, this corresponds to the (ugly) \texttt{<DI::Atom as Nature>::RV<2>} type.
    \item \texttt{DI\_Atom\_RV2} implements trait \rustdoc{measurements/noise_threshold/trait}{MakeNoiseThreshold}.
        That is, it must be possible to build the mechanism from this new equivalent distribution.
    \item type \texttt{(DI, MI)} implements trait \rustdoc{core/trait}{MetricSpace}
\end{itemize}

\subsubsection*{User-Verified}
None

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/MakeNoiseThreshold_for_DiscreteLaplace.py}

\subsection*{Postcondition}
\begin{theorem}
    \validMeasurement{(\texttt{self, input\_space, threshold, DI, MI, MO})}{\texttt{make\_noise\_threshold}}
\end{theorem}

\begin{proof}
    On line \ref{line:make-noise}, \texttt{make\_noise\_threshold} has no preconditions, 
    so irregardless of any prior logic,
    the postcondition of \texttt{make\_noise\_threshold} follows that the output is a valid measurement.
\end{proof}

The complexity in the type system here is designed to be free of privacy implications,
to help simplify the core, privacy-sensitive implementation.

\end{document}
