\documentclass{article}
\input{../../lib.sty}

\title{\texttt{fn make\_row\_by\_row}}
\author{Michael Shoemate}
\date{}

\begin{document}

\maketitle

\contrib
Proves soundness of \rustdoc{transformations/fn}{make\_row\_by\_row} in \asOfCommit{mod.rs}{f5bb719}.
This constructor is a special case of \rustdoc{transformations/fn}{make\_row\_by\_row\_fallible}.
The proof for this constructor appeals to the proof for \rustdoc{transformations/fn}{make\_row\_by\_row\_fallible}.

\texttt{make\_row\_by\_row} returns a Transformation that applies a user-specified function to each record in the input dataset.

\subsection*{Vetting History}
\begin{itemize}
    \item \vettingPR{562}
\end{itemize}

\section{Hoare Triple}
\subsection*{Precondition}
\begin{itemize}

    \item \texttt{DI} (input domain) is a type with trait \rustdoc{transformations/trait}{RowByRowDomain}\texttt{<DO>}. 
        This trait provides a way to apply a map function to each record in the input dataset to retrieve a dataset that is a member of the output domain, of type \texttt{DO}. The trait further implies that \texttt{DatasetDomain} is also implemented for \texttt{DI}.
    \item \texttt{DO} (output domain) is a type with trait \rustdoc{transformations/trait}{DatasetDomain}. 
        \texttt{DatasetDomain} is used to define the type of the row domain.
    \item \texttt{M} (metric) is a type with trait \rustdoc{transformations/trait}{DatasetMetric}. 
        \texttt{DatasetMetric} is used to restrict the set of valid metrics to those which measure distances between datasets.
    \item \rustdoc{core/trait}{MetricSpace} is implemented for \texttt{(DI, M)}. Therefore \texttt{M} is a valid metric on \texttt{DI}.
    \item \rustdoc{core/trait}{MetricSpace} is implemented for \texttt{(DO, M)}.
    \item \texttt{row\_function} has no side-effects.
    \item If the input to \texttt{row\_function} is a member of \texttt{input\_domain}'s row domain, then the output is a member of \texttt{output\_row\_domain}.
\end{itemize}

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/make_row_by_row.py}

\subsection*{Postcondition}
\validTransformation{\texttt{(input\_domain, input\_metric, output\_domain, row\_function, DI, DO, M)}}{\texttt{make\_row\_by\_row}}

\section{Proofs}

\begin{proof} \textbf{(Part 1 -- appropriate output domain).}
    Since the preconditions for this constructor are a super-set of the preconditions on \texttt{make\_row\_by\_row\_fallible}, 
    the proof of \texttt{make\_row\_by\_row\_fallible} applies. 
    Thus, by the output domain proof on \texttt{make\_row\_by\_row\_fallible}, 
    for all settings of input arguments, the function returns a dataset in the output domain.
\end{proof}

\begin{proof} \textbf{(Part 2 -- stability map).} 
    The proof of \texttt{make\_row\_by\_row\_fallible} similarly applies. 
    Thus, by the stability map proof on \texttt{make\_row\_by\_row\_fallible}, 
    for all settings of input arguments, 
    where $u, v$ are \din-close under \texttt{input\_metric},
    \function(u), \function(v) are \dout-close under \texttt{output\_metric}.
\end{proof}

\end{document}
