\documentclass{article}
\input{../../lib.sty}

\title{\texttt{fn make\_is\_equal}}
\author{S\'ilvia Casacuberta}
\date{}

\begin{document}

\maketitle

\contrib
Proves soundness of \texttt{fn make\_is\_equal} in \asOfCommit{mod.rs}{0db9c6036}.
The transformation checks if each element in a vector dataset is equivalent to a given value,
returning a vector of booleans.

\subsection*{Vetting History}
\begin{itemize}
    \item \vettingPR{512}
\end{itemize}

\section{Hoare Triple}
\subsection*{Precondition}
To ensure the correctness of the output, we require the following preconditions:

\begin{itemize}
    \item Type \texttt{TIA} must have trait \texttt{PartialEq}.
    \item Type \texttt{M} must have trait \texttt{DatasetMetric}.
    \item \rustdoc{core/trait}{MetricSpace} is implemented for \texttt{(VectorDomain<AtomDomain<TIA>{}>, M)}. Therefore \texttt{M} is a valid metric on \texttt{VectorDomain<AtomDomain<TIA>{}>}.
    \item \rustdoc{core/trait}{MetricSpace} is implemented for \texttt{(VectorDomain<AtomDomain<bool>{}>, M)}.
\end{itemize}

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/make_is_equal.py}

\subsubsection*{Postconditions}
\validTransformation{\texttt{(input\_domain, input\_metric, value)}}{\texttt{make\_is\_equal}}

\section{Proof}

\begin{lemma}
    The invocation of \rustdoc{transformations/manipulation/fn}{make\_row\_by\_row} (line \ref{line:row-by-row}) satisfies its preconditions.
\end{lemma}

\begin{proof}
    \label{lemma:row-by-row-precondition}
    The preconditions of \texttt{make\_is\_equal} and pseudocode definition (line \ref{line:def}) ensure that the type preconditions of \texttt{make\_row\_by\_row} are satisfied. 
    The remaining preconditions of \texttt{make\_row\_by\_row} are:
    \begin{itemize}
        \item \texttt{row\_function} has no side-effects.
        \item If the input to \texttt{row\_function} is a member of \texttt{input\_domain}'s row domain, then the output is a member of \texttt{output\_row\_domain}.
    \end{itemize}

    The first precondition is satisfied by the definition of \texttt{is\_equal} (line \ref{line:function}) in the pseudocode.

    For the second precondition, assume the input is a member of \texttt{input\_domain}'s row domain. 
    By the definition of \texttt{PartialEq}, the output of \texttt{is\_equal} is boolean, and \texttt{AtomDomain<bool>} includes all booleans.
    Therefore, the output is a member of \texttt{output\_row\_domain}.
\end{proof}

We now prove the postcondition of \texttt{make\_is\_equal}.
\begin{proof}
By \ref{lemma:row-by-row-precondition}, the preconditions of \rustdoc{transformations/manipulation/fn}{make\_row\_by\_row} are satisfied.
Thus, by the definition of \\\texttt{make\_row\_by\_row}, the output is a valid transformation.
\end{proof}

\end{document}