\documentclass{article}
\input{../../lib.sty}

\title{\texttt{impl PrivateCandidateMeasure for RenyiDivergence}}
\author{Michael Shoemate}
\date{}

\begin{document}
\maketitle

Proves soundness of the \texttt{RenyiDivergence} implementation of
\texttt{PrivateCandidateMeasure} in \texttt{mod.rs}.

\section{Hoare Triple}
\subsection*{Precondition}
\texttt{base\_map} is the privacy map of a valid \texttt{RenyiDivergence} measurement.

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/renyi_divergence_private_candidate_measure.py}

\subsection*{Postcondition}
\begin{theorem}
\label{postcondition}
The \texttt{validate} method accepts exactly the supported Rényi-DP combinations implemented in
\texttt{mod.rs}, and \texttt{new\_privacy\_map} returns the privacy map stated in the proof
definition of \texttt{make\_select\_private\_candidate}.
\end{theorem}

\section{Proof}
Validation is immediate: line \ref{line:geometric-guard} rejects thresholded constructions unless
the repetition family is geometric, and no other combinations are rejected.

For privacy accounting, line \ref{line:base-curve} evaluates the base privacy map to obtain the
base Rényi curve.
There are three accepted branches.
\begin{itemize}
    \item In the thresholded geometric case, line \ref{line:conditional-branch} returns
    \rustdoc{combinators/select\_private\_candidate/bounds/fn}{new\_conditional\_rdp\_curve}.
    By its postcondition, this upper-bounds the Appendix C / Corollary 16
    conditional-sampling curve.
    \item In the non-thresholded negative-binomial case, line \ref{line:nb-branch} returns
    \rustdoc{combinators/select\_private\_candidate/bounds/fn}{new\_negative\_binomial\_rdp\_curve}.
    By its postcondition, this upper-bounds the Theorem 2 curve.
    \item In the non-thresholded Poisson case, line \ref{line:poisson-branch} returns
    \rustdoc{combinators/select\_private\_candidate/bounds/fn}{new\_poisson\_rdp\_curve}. By its
    postcondition, this upper-bounds the implemented Theorem 6-based curve.
\end{itemize}

Therefore the returned privacy map is sound in every accepted branch.

\end{document}
