\documentclass{article} 
\input{../../../lib.sty} 
 
\title{\texttt{fn sample\_bernoulli\_exp1}} 
\author{Michael Shoemate} 
 
\begin{document} 
\maketitle 
 
Proves soundness of \texttt{fn sample\_bernoulli\_exp1} in \asOfCommit{mod.rs}{0be3ab3e6}.\\ 
\texttt{fn sample\_bernoulli\_exp1} returns a sample from the $Bernoulli(exp(-x))$ distribution for some rational argument in $[0, 1]$. 
This proof is adapted from \href{https://arxiv.org/pdf/2004.00010.pdf#subsection.5.1}{subsection 5.1} of \cite{CKS20}. 
 
\section{Hoare Triple} 
\subsection*{Preconditions} 
 
\textbf{Compiler-verified} 
\begin{itemize} 
    \item Argument $\texttt{x}$ is of type \texttt{RBig}, a rational bignum 
\end{itemize} 
 
\textbf{User-verified} 
$x \in [0, 1]$ 
 
 
\subsection*{Pseudocode}         
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/sample_bernoulli_exp1.py} 
 
\subsection*{Postcondition} 
\begin{theorem}
\label{postcondition} 
For any setting of the input parameter \texttt{x} such that the given preconditions hold, \\ 
\texttt{sample\_bernoulli\_exp1} either returns \texttt{Err(e)} due to a lack of system entropy, 
or \texttt{Ok(out)}, where \texttt{out} is distributed as $Bernoulli(exp(-x))$. 
\end{theorem}
 
\section{Proof} 
Assume the preconditions are met. 
 
\begin{lemma} 
    \label{err-e} 
    \texttt{sample\_bernoulli\_exp1} only returns \texttt{Err(e)} when there is a lack of system entropy. 
\end{lemma} 
 
\begin{proof} 
    In all usages of \rustdoc{traits/samplers/bernoulli/fn}{sample\_bernoulli\_rational}, the argument passed satisfies its definition preconditions,  
    by the preconditions on \texttt{x} and function logic. 
    Thus, by its postcondition, \texttt{sample\_bernoulli\_rational} only returns an error when there is a lack of system entropy. 
    The only source of errors in \texttt{sample\_bernoulli\_exp1} is from the invocation of \texttt{sample\_bernoulli\_rational}. 
    Therefore \texttt{sample\_bernoulli\_exp1} only returns \texttt{Err(e)} when there is a lack of system entropy. 
\end{proof} 
 
\begin{lemma} 
    \label{P_K_gt_n} 
    Let $K^{*}$ denote the final value of \texttt{k} on line \ref{line:K}. 
    Then $P[K^{*} > n] = \frac{x^n}{n!}$ for any integer $n \ge 0$ \cite{CKS20}. 
\end{lemma} 
 
\begin{proof} 
    For $k  > 0$, let $a_k$ denote the $k^{th}$ outcome of \texttt{sample\_bernoulli\_rational} on line \ref{line:B_i}. 
    By the definition of \texttt{sample\_bernoulli\_rational}, under the established conditions and preconditions, 
    each $A_k$ is distributed as $Bernoulli(x / k)$. 
 
 If $n \ge 1$, we have:
    \begin{align*} 
        P[K^{*} > n] &= P[A_1 = A_2 = ... = A_n = \top] && \text{since } K^{*} > n \iff \forall k \leq n, a_k = \top \\ 
        &= \prod_{k=1}^n P[A_k = \top] && \text{all $A_k$ are independent} \\ 
        &= \prod_{k=1}^n \frac{x}{k} && \text{since $A_k \sim Bernoulli(x/k)$} \\ 
        &= \frac{x^n}{n!}
    \end{align*} 
  If $n = 0$, we also have $P[K^{*} > 0] = 1 = \frac{x^0}{0!}$.
\end{proof} 
 
\begin{lemma} 
    \label{ok-out} 
    $is\_odd(K^{*}) \sim Bernoulli(exp(-x))$ \cite{CKS20}. 
\end{lemma} 
 
\begin{proof} 
    \begin{align*} 
        P[K^{*} \text{ odd}] &= \sum_{k=0}^\infty P[K^{*} = 2k + 1] \\ 
        &= \sum_{k=0}^\infty (P[K^{*} > 2k] - P[K^{*} > 2k + 1]) \\ 
        &= \sum_{k=0}^\infty \left(\frac{x^{2k}}{(2k)!} - \frac{x^{2k + 1}}{(2k + 1)!}\right) && \text{by } \ref{P_K_gt_n} \\ 
        &= exp(-x) 
    \end{align*} 
 
\end{proof} 
 
\begin{proof} 
    Since $\mathtt{k}$ is distributed according to $K^{*}$, then by \ref{ok-out}, \texttt{out} is distributed as $Bernoulli(exp(-x))$. 
    Together with \ref{err-e}, Theorem~\ref{postcondition} holds.
\end{proof} 
 
\bibliographystyle{alpha} 
\bibliography{mod} 
\end{document}