\documentclass{article}
\input{../../../lib.sty}

\title{\texttt{fn sample\_discrete\_gaussian}}
\author{Michael Shoemate}

\begin{document}
\maketitle

\contrib
Proves soundness of \texttt{fn sample\_discrete\_gaussian} in \asOfCommit{mod.rs}{0be3ab3e6}.
This proof is an adaptation of \href{https://arxiv.org/pdf/2004.00010.pdf#subsection.5.3}{subsection 5.3} of \cite{CKS20}.

\subsection*{Vetting history}
\begin{itemize}
    \item \vettingPR{519}
\end{itemize}

\section{Hoare Triple}
\subsection*{Precondition}
$\texttt{scale} \in \mathbb{Q} \land \texttt{scale} \geq 0$

\subsection*{Implementation}        
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/sample_discrete_gaussian.py}

\subsection*{Postcondition}
\label{postcondition}
For any setting of the input parameter \texttt{scale} such that the given preconditions hold, \\
\texttt{sample\_discrete\_gaussian} either returns \texttt{Err(e)} due to a lack of system entropy,
or \texttt{Ok(out)}, where \texttt{out} is distributed as $\mathcal{N}_\mathbb{Z}(0, scale^2)$.

\section{Proof}

\begin{definition}
    (Discrete Gaussian). \cite{CKS20} Let $\mu, \sigma \in \mathbb{R}$ with $\sigma > 0$. 
    The discrete gaussian distribution with location $\mu$ and scale $\sigma$ is denoted $\mathcal{N}_\mathbb{Z}(\mu, \sigma^2)$. 
    It is a probability distribution supported on the integers and defined by
    \begin{equation*}
        \forall x \in \mathbb{Z} \quad  P[X = x] = \frac{e^{-\frac{(x - \mu)^2}{2\sigma^2}}}{\sum_{y\in\mathbb{Z}}e^{-\frac{(y - \mu)^2}{2\sigma^2}}} \quad \text{where } X \sim \mathcal{N}_\mathbb{Z}(\mu, \sigma^2)
    \end{equation*}
\end{definition}


\begin{lemma}
    \label{err-e}
    \texttt{sample\_discrete\_gaussian} only returns \texttt{Err(e)} when there is a lack of system entropy.
\end{lemma}

\begin{proof}
    By the non-negativity precondition on \texttt{scale}, \texttt{t} on line \ref{line:t} is non-negative, 
    so the precondition on \rustdoc{traits/samplers/cks20/fn}{sample\_discrete\_laplace} is met.
    Similarly, since \texttt{bias} on line \ref{line:bias} is non-negative, the preconditions on \rustdoc{traits/samplers/cks20/fn}{sample\_bernoulli\_exp} are met.
    By the definitions of \texttt{sample\_discrete\_laplace} and \texttt{sample\_bernoulli\_exp}, an error is only returned when there is a lack of system entropy.
    The only source of errors in \texttt{sample\_discrete\_gaussian} is from the invocation of these functions,
    therefore \texttt{sample\_discrete\_gaussian} only returns \texttt{Err(e)} when there is a lack of system entropy.
\end{proof}

We now condition on not returning an error.
Let $t = \lfloor \sigma \rfloor + 1$, and fix any iteration of the loop. 

\begin{lemma}\cite{CKS20}
    \label{E_C}
    If $y$ is a realization of $Y \sim \mathcal{L}_\mathbb{Z}(0, t)$, and $c$ is a realization of $C \sim Bernoulli(exp(-(|y| - \sigma^2 / t)^2 / (2 \sigma^2)))$, then
    $E[C] = \frac{1 - e^{-1/\sigma}}{1 + e^{-1/\sigma}}e^{-\frac{\sigma^2}{2t^2}} \sum_{y\in \mathbb{Z}} e^{-\frac{y^2}{2\sigma^2}}$.
\end{lemma}

\begin{proof}
    \begin{align*}
        E[C] &= E[E[C|Y]] \\
        &= E[e^{-\frac{(|Y| - \sigma^2/t)^2}{2\sigma^2}}] && \text{since } E[Bernoulli(p)] = p \\
        &= \frac{1 - e^{-1/\sigma}}{1 + e^{-1/\sigma}} \sum_{y\in \mathbb{Z}} e^{-\frac{(|y| - \sigma^2/t)^2}{2\sigma^2} - |y|/t} && \text{expectation over } Y \sim \mathcal{L}_\mathcal{Z}(0, \sigma) \\
        &= \frac{1 - e^{-1/\sigma}}{1 + e^{-1/\sigma}}e^{-\frac{\sigma^2}{2t^2}} \sum_{y\in \mathbb{Z}} e^{-\frac{y^2}{2\sigma^2}}
    \end{align*}
\end{proof}

We now show that conditioning $Y$ on the success of $C$ gives the desired output distribution.
\begin{theorem}\cite{CKS20}
    \label{P_Yy_CT} If $y$ is a realization of $Y \sim \mathcal{L}_\mathbb{Z}(0, t)$ and $c$ is a realization of $C \sim Bernoulli(exp(-(|y| - \sigma^2 / t)^2 / (2 \sigma^2)))$, then
    $P[Y=y | C=\top] = \frac{e^{-\frac{y^2}{2\sigma^2}}}{\sum_{y' \in \mathbb{Z}} e^{-\frac{y'^2}{2\sigma^2}}}$. That is, $Y|_{C=\top} \sim \mathcal{N}_\mathbb{Z}(0, \sigma^2)$.
\end{theorem}

\begin{proof}
    \begin{align*}
        P[Y=y | C=\top] &= \frac{P[C=\top|Y=y]P[Y=y]}{P[C=\top]} && \text{Bayes' Theorem} \\
        &= \frac{e^-\frac{(|y| - \sigma^2/t)^2}{2\sigma^2} \frac{1 - e^{-1/t}}{1 + e^{-1/t}} e^{-|y|/t}}{E[C]} && \text{by definition of } \mathcal{L}_\mathbb{Z}(0, \sigma) \\
        &= \frac{e^-\frac{(|y| - \sigma^2/t)^2}{2\sigma^2} e^{-|y|/t}}{e^{-(\sigma/t)^2/2} \sum_{y' \in \mathbb{Z}} e^{-\frac{y'^2}{2\sigma^2}}} && \text{by } \ref{E_C}\\
        &= \frac{e^{-\frac{y^2}{2\sigma^2}}}{\sum_{y' \in \mathbb{Z}} e^{-\frac{y'^2}{2\sigma^2}}}
    \end{align*}
\end{proof}

\begin{lemma}\label{ok-out}
    If the outcome of \texttt{sample\_discrete\_gaussian} is \texttt{Ok(out)}, 
    then \texttt{out} is distributed as $\mathcal{N}_\mathbb{Z}(0, scale^2)$.
\end{lemma}

\begin{proof}
    In the \ref{err-e} proof, it was established that the preconditions on \texttt{sample\_discrete\_laplace} are met, 
    so \texttt{candidate} on line \ref{line:candidate} is distributed as $\mathcal{L}_\mathbb{Z}(0, t)$.
    Similarly, by the definition of \texttt{sample\_bernoulli\_exp}, the outcome is distributed according to $Bernoulli(exp(-(|y| - \sigma^2 / t)^2 / (2 \sigma^2)))$.
    Since on line \ref{line:bern}, we condition returning \texttt{candidate} on a $\top$ sample, 
    the conditions to apply \ref{P_Yy_CT} are met. 
    Therefore \texttt{out} is distributed as $\mathcal{N}_\mathbb{Z}(0, scale^2)$.
\end{proof}

\begin{proof}
    \ref{postcondition} holds by \ref{err-e} and \ref{ok-out}.
\end{proof}

\bibliographystyle{alpha}
\bibliography{mod}

\end{document}