\documentclass{article}
\input{../../../lib.sty}

\title{\texttt{fn sample\_discrete\_laplace}}
\author{Michael Shoemate}

\begin{document}
\maketitle

\contrib
Proves soundness of \texttt{fn sample\_discrete\_laplace} in \asOfCommit{mod.rs}{0be3ab3e6}.
This proof is an adaptation of \href{https://arxiv.org/pdf/2004.00010.pdf#subsection.5.2}{subsection 5.2} of \cite{CKS20}.

\subsection*{Vetting history}
\begin{itemize}
    \item \vettingPR{519}
    \item \vettingPR{1134}
\end{itemize}

\section{Hoare Triple}
\subsection*{Precondition}
$\texttt{scale} \in \mathbb{Q} \land \texttt{scale} > 0$

\subsection*{Pseudocode}        
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/sample_discrete_laplace.py}

\subsection*{Postcondition}
\label{postcondition}
For any setting of the input parameter \texttt{scale} such that the given preconditions hold, \\
\texttt{sample\_discrete\_laplace} either returns \texttt{Err(e)} due to a lack of system entropy,
or \texttt{Ok(out)}, where \texttt{out} is distributed as $\mathcal{L}_\mathbb{Z}(0, scale)$.

\section{Proof}
\begin{definition} \cite{BV17}
    (Discrete Laplace). Let $\mu, \sigma \in \mathbb{R}$ with $\sigma > 0$. 
    The discrete laplace distribution with location $\mu$ and scale $s$ is denoted $\mathcal{L}_\mathbb{Z}(\mu, s)$. 
    It is a probability distribution supported on the integers and defined by
    \begin{equation*}
        \forall x \in \mathbb{Z} \quad  P[X = x] = \frac{e^{1/s} - 1}{e^{1/s} + 1} e^{-|x|/s} \quad \text{where } X \sim \mathcal{L}_\mathbb{Z}(\mu, s)
    \end{equation*}
\end{definition}

Assume the preconditions are met.

\begin{lemma}
    \label{err-e}
    \texttt{sample\_discrete\_laplace} only returns \texttt{Err(e)} when there is a lack of system entropy.
\end{lemma}

\begin{proof}
    By the non-negativity precondition on \texttt{scale},
    the precondition on \rustdoc{traits/samplers/cks20/fn}{sample\_geometric\_exp\_fast} is met.
    By the definitions of \texttt{sample\_geometric\_exp\_fast} and \texttt{sample\_standard\_bernoulli}, 
    an error is only returned when there is a lack of system entropy.
    The only source of errors is from the invocation of these functions,
    therefore \texttt{sample\_discrete\_gaussian} only returns \texttt{Err(e)} when there is a lack of system entropy.
\end{proof}

We now condition on not returning an error, and establish some helpful lemmas.
\begin{lemma}\cite{CKS20}\label{P_B_Y_ne_T_0}
    Let $B \sim Bernoulli(1/2)$ and $Y \sim Geometric(1 - e^{-1/s})$ for some $s > 0$. 
    Then $P[(B, Y) \neq (\top, 0)] = \frac{1}{2} (e^{-1/s} + 1)$.
\end{lemma}

\begin{proof}
    \begin{align*}
        P[(B, Y) \neq (\top, 0)] &= P[B = \top, Y > 0] + P[B = \bot] && \text{by LOTP} \\
        &= P[B = \top] P[Y > 0] + P[B = \bot] && \text{by independence of B, Y} \\
        &= \frac{1}{2} e^{-1/s} + \frac{1}{2} \\
        &= \frac{1}{2} (e^{-1/s} + 1)
    \end{align*}
\end{proof}

\begin{lemma}\cite{CKS20}\label{P_Lx_BY_ne_T0}
    Given random variables $B \sim Bernoulli(1/2)$ and $Y \sim Geometric(1 - e^{-1/s})$, define $X|_{B=\top} = Y$, and $X|_{B=\bot} = -Y$. 
    If $(B, Y) \neq (\top, 0)$, then $X \sim \mathcal{L}_\mathbb{Z}(0, scale)$. 
    That is, $P[X = x | (B, Y) \neq (\top, 0)] = \frac{e^{1/s} - 1}{e^{1/s} + 1} e^{-|x|/s}$ for any $x \in \mathbb{Z}$.
\end{lemma}

\begin{proof}
\begin{align*}
P[X = x | (B, Y) \neq (\top, 0)] &= \frac{P[X = x, (B, Y) \neq (\top, 0)]}{P[(B, Y) \neq (\top, 0)]} \\
    &= \frac{P[X = |x|, B = \mathbb{I}[x < 0]]}{P[(B, Y) \neq (\top, 0)]} && \text{since x = }\pm y \\
    &= \frac{P[X = |x|] P[B = \mathbb{I}[x < 0]]}{P[(B, Y) \neq (\top, 0)]} && \text{by independence of B, Y} \\
    &= \frac{P[X = |x|] \frac{1}{2}}{\frac{1}{2} (e^{-1/s} + 1)} && \text{by } \ref{P_B_Y_ne_T_0} \\
    &= \frac{1 - e^{-1/s}}{1 + e^{-1/s}} e^{-|x|/s} \\
    &= \frac{e^{1/s} - 1}{e^{1/s} + 1} e^{-|x|/s} \\
\end{align*}
\end{proof}

\begin{lemma}\label{ok-out}
    If the outcome of \texttt{sample\_discrete\_laplace} is \texttt{Ok(out)}, 
    then \texttt{out} is distributed as $\mathcal{L}_\mathbb{Z}(0, scale)$.
\end{lemma}

\begin{proof}
    In the \ref{err-e} proof, it was established that the preconditions on \texttt{sample\_geometric\_exp\_fast} are met.
    therefore \texttt{magnitude} on line \ref{line:magnitude} is distributed as $ Geometric(1 - e^{-1/scale})$.
    Similarly, by the definition of \texttt{sample\_standard\_bernoulli}, \texttt{sign} is distributed according to $Bernoulli(p=1/2)$.
    The branching logic from line \ref{line:branching} on satisfies the procedures described in \ref{P_Lx_BY_ne_T0}.
    Therefore, by \ref{P_Lx_BY_ne_T0}, \texttt{out} is distributed as $\mathcal{L}_\mathbb{Z}(0, scale)$.
\end{proof}

\begin{proof}
    \ref{postcondition} holds by \ref{err-e} and \ref{ok-out}.
\end{proof}


\bibliographystyle{alpha}
\bibliography{mod}

\end{document}