\documentclass{article}
\input{../../../../lib.sty}

\title{\texttt{MakeNoise<VectorDomain<AtomDomain<T>>, LpDistance<P, QI>, MO> for FloatExpFamily<P>}}
\author{Michael Shoemate}
\date{}

\begin{document}

\maketitle

\contrib
Proves soundness of the implementation of \rustdoc{measurements/noise/trait}{MakeNoise} over vectors
for \texttt{FloatExpFamily} in \asOfCommit{mod.rs}{f5bb719}.

This mechanism samples from the \texttt{FloatExpFamily} distribution, 
where the tails saturate to the positive and negative infinity floats in the native float type.
This is done by rounding floats to a fractional grid, 
adding noise from from the \texttt{ZExpFamily} distribution (supported on all integers) 
to the numerator of the fraction (with noise scale multiplied by the denominator),
and then converting back to the nearest float, or saturating to positive or negative infinity.

\section{Hoare Triple}
\subsection*{Precondition}
\subsubsection*{Compiler-Verified}

\begin{itemize}
    \item Generic \texttt{T} implements trait \rustdoc{traits/trait}{Float}
    \item Const-generic \texttt{P} is of type \texttt{usize}
    \item Generic \texttt{QI} implements trait \rustdoc{traits/trait}{Number}
    \item Generic \texttt{MO} implements trait \rustdoc{core/trait}{Measure}
    \item Type \texttt{i32} implements trait \rustdoc{traits/trait}{ExactIntCast}\texttt{<<T as FloatBits>::Bits>},
        This requirement means that the raw bits of \texttt{T} can be exactly cast to an \texttt{i32}.
    \item Type \texttt{RBig} implements traits \texttt{TryFrom<T>} and \texttt{TryFrom<QI>}.
        This is for fallible exact casting from input sensitivity to a rational in the privacy map.
    \item Type \texttt{ZExpFamily<P>} implements traits \rustdoc{measurements/trait}{NoisePrivacyMap}\texttt{<LpDistance<P, RBig>, MO>}
        This bound requires that it must be possible to construct a privacy map for this combination of noise distribution, distance type and privacy measure.
\end{itemize}

\subsubsection*{User-Verified}
None

\subsection*{Pseudocode}
\lstinputlisting[language=Python,firstline=2,escapechar=|]{./pseudocode/MakeNoise_VectorDomain_for_FloatExpFamily.py}

\subsection*{Postcondition}
\begin{theorem}
    \validMeasurement{(\texttt{self, input\_space, MO, T, P, QI})}{\texttt{make\_noise}}
\end{theorem}

\begin{proof}
    Line \ref{line:dist} constructs a new random variable following a distribution equivalent to \texttt{FloatExpFamily}, 
    but without tails that saturate to infinity, and with gaps between adjacent point on the grid adjusted to one.

    Neither constructor \rustdoc{measurements/noise/nature/float/fn}{make\_float\_to\_bigint} nor
    \rustdoc{measurements/noise/trait}{MakeNoise}\texttt{.make\_noise} have manual preconditions, 
    and the postconditions guarantee a valid transformation and valid measurement, respectively.
    \rustdoc{measurements/noise/nature/float/fn}{then\_deintegerize\_vec} also does not have preconditions,
    and its postcondition guarantees that it returns a valid postprocessor.

    The chain of a valid transformation, valid measurement and valid postprocessor is a valid measurement.
\end{proof}

\end{document}
